A Large and Variable Leading Tail of Helium in a Sub-Saturn Undergoing Runaway Inflation

### 2023-07-12 paper/paper1/ms.tex

\begin{abstract}
    Atmospheric escape shapes the fate of exoplanets, with statistical evidence for transformative mass loss imprinted across the mass-radius-insolation distribution. The Habitable Zone Planet Finder (HPF) Spectrograph Exospheres program aims to expand the search across a diverse range of planet and host star properties via the 10833 \AA~ Helium triplet, a sensitive probe of active atmospheric escape. Here we present a detection of 10\% absorption depth during transit spectroscopy of the highly irradiated, low-gravity, inflated hot Saturn HAT-P-67~b. The 13.8 hours of on-sky integration time over 39 nights sample the entire planet orbit, uncovering excess Helium absorption preceding the transit by up to 130 planetary radii in a large leading tail. This configuration can be understood as the escaping material overflowing its small Roche lobe, and advecting most of the gas into the stellar—and not planetary---rest frame, consistent with the Doppler velocity structure seen in the Helium line profiles. The prominent leading tail serves as direct evidence for dayside mass loss with a strong day-/night- side asymmetry.  We see some transit-to-transit variability in the line profile, consistent with the interplay of stellar and planetary winds.  We employ 1D Parker wind models to estimate the mass loss rate, finding values on the order of $2\times10^{13}$ g/s, with large uncertainties owing to the unknown XUV flux of the F host star.  The large mass loss in HAT-P-67~b represents a valuable example of an inflated hot Saturn, a class of planets recently identified to be rare as their atmospheres are predicted to evaporate quickly.  We contrast two physical mechanisms for runaway evaporation: Ohmic dissipation and XUV irradiation, slightly favoring the latter.   
\end{abstract}

@@

Uncertainty in system ages, evaporation timescales, X-ray/UV radiation, and dominating physical mechanisms degrade our ability to foretell if any given planet will exhibit ongoing signatures of atmospheric escape.  Episodic stellar wind gusts and other forms of astrophysical variability could also subdue the appearance of atmospheric escape, even where we expect it most.  The 48 published non-detections of atmospheric escape \citep{2022arXiv221116243D} must encode these natural whims in a way that we have not yet disentangled. Nevertheless, we can boost our chances of witnessing active and significant atmospheric escape by targeting sources that seem predisposed to loss.  These intrinsic or extrinsic factors may include proximity to the host star, low surface gravity, and high energy incident radiation.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/HAT-P-67b_radius_valley.png}
    \caption{Overview of exoplanet demographics.  The pixel bins reflect the observed density of over 5000 planets accessed from the NASA Exoplanet Archive. Detections of atmospheric escape are common among large planets with strong insolation.  \hatpb---among the most inflated planets known---is an outlier.}
    \label{fig:instellation}
\end{figure}

\begin{figure}
    \includegraphics[width=\linewidth]{figures/tf2018_fig2_update2023_HAT.png}
    \caption{Mass-radius trends for inflated hot sub-Saturns and hot Jupiters, form modeled after Figure 2 from \citet{2018AJ....155..214T} and updated with NASA Exoplanet Archive confirmed planets.  The trend lines show the mass-radius relationship for equilibrium temperatures of 500, 1000, 1250, 1500, and 2000 K, assuming the mean composition and mean heating efficiency from the original figure.  HAT-P-67~b stands alone in a region defined by the lack of inflated sub-Saturns and explained by short lifetimes from runaway inflation and mass loss.}
    \label{fig:ThornFortUpdated}
\end{figure}

@@

The TESS Sector 26 lightcurve exhibits a weak $\sim$3 ppt peak-to-valley out-of-transit modulation. TESS Sectors 51-53 do not show as conspicuous a modulation signal but still show some ostensibly stellar variability.  The ebb and flow of the modulation can be seen in the minimally processed TESS lightcurve in Figure \ref{fig:TESSoverview}.  We fit the TESS lightcurve modulation with a quasiperiodic Gaussian Process (GP) model using \texttt{celerite} \citep{celerite1,celerite2}.  The GP model fit to the entire composite lightcurve yields a 4.7-day period; when fit to individual sectors alone the periods hover around 5.9 days.

We can constrain the expected stellar rotation period based on the stellar radius estimate, $v\sin{i_\star}$, and the observation of the planet's orbital inclination $i_p\sim90^\circ$ from orbit fitting and Doppler tomography \citep{2017AJ....153..211Z}.  We assume spin-orbit alignment, $i_\star \sim i_p$.  We adopt a high limit of $v\sin{i}=35.8\pm1.1$ km/s and a lower $v\sin{i}=30.9\pm2$ km/s value if macroturbulence is accounted for.  We obtain a range of $3.2 < P_\mathrm{rot}  < 4.8 $ days.  This range is typical for F stars that have not yet evolved too far into the subgiant branch \citep{2022ApJ...930....7A}.  The geometrical constraint comports with the 4.7-day GP-based modulation period derived from the stitched TESS lightcurve, but is lower than the 5.9-day per-Sector fits, slightly preferring the lower 4.7-day modulation as the stellar rotation period.  The 5.9-day value would require a stellar radius over 3.2$R_\odot$, which appears implausible.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/phase_2D_diagram_resid.png}
    \caption{Absorption depth phase scan over the entire HAT-P-67b orbit from 41 visits with HPF.  The planet orbital rest frame is shown as the black sine wave. Gaps in data at 10834-10836 \AA arise from telluric masking.  Unmasked telluric lines are perceptible at 10825 \AA.}
    \label{fig:HPFscanResid}
\end{figure}

>>>>

\begin{abstract}
    Atmospheric escape shapes the fate of exoplanets, with statistical evidence for transformative mass loss imprinted across the mass-radius-insolation distribution. The Habitable Zone Planet Finder (HPF) Spectrograph Exospheres program aims to expand the search across a diverse range of planet and host star properties via the 10833 \AA~ Helium triplet, a sensitive probe of active atmospheric escape. Here we present a detection of up to 10\% absorption depth during transit spectroscopy of the highly irradiated, low-gravity, inflated hot Saturn HAT-P-67~b. The 13.8 hours of on-sky integration time over 39 nights sample the entire planet orbit, uncovering excess Helium absorption preceding the transit by up to 130 planetary radii in a large leading tail. This configuration can be understood as the escaping material overflowing its small Roche lobe, and advecting most of the gas into the stellar—and not planetary---rest frame, consistent with the Doppler velocity structure seen in the Helium line profiles. The prominent leading tail serves as direct evidence for dayside mass loss with a strong day-/night- side asymmetry.  We see some transit-to-transit variability in the line profile, consistent with the interplay of stellar and planetary winds.  We employ 1D Parker wind models to estimate the mass loss rate, finding values on the order of $2\times10^{13}$ g/s, with large uncertainties owing to the unknown XUV flux of the F host star.  The large mass loss in HAT-P-67~b represents a valuable example of an inflated hot Saturn, a class of planets recently identified to be rare as their atmospheres are predicted to evaporate quickly.  We contrast two physical mechanisms for runaway evaporation: Ohmic dissipation and XUV irradiation, slightly favoring the latter.   
\end{abstract}

@@

Uncertainty in system ages, evaporation timescales, X-ray/UV radiation, and dominating physical mechanisms degrade our ability to foretell if any given planet will exhibit ongoing signatures of atmospheric escape.  Episodic stellar wind gusts and other forms of astrophysical variability could also subdue the appearance of atmospheric escape, even where we expect it most.  The over 57 published non-detections of atmospheric escape \citep{2022arXiv221116243D,2023arXiv230700967G} must encode these natural whims in a way that we have not yet disentangled. Nevertheless, we can boost our chances of witnessing active and significant atmospheric escape by targeting sources that seem predisposed to loss.  These intrinsic or extrinsic factors may include proximity to the host star, low surface gravity, and high energy incident radiation.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/HAT-P-67b_radius_valley.png}
    \caption{Overview of exoplanet demographics.  The pixel bins reflect the observed density of over 5000 planets accessed from the NASA Exoplanet Archive. Detections of atmospheric escape are common among large planets with strong insolation.  \hatpb is among the most inflated planets known.}
    \label{fig:instellation}
\end{figure}

\begin{figure}
    \includegraphics[width=\linewidth]{figures/tf2018_fig2_update2023_HAT.png}
    \caption{Mass-radius trends for inflated hot sub-Saturns and hot Jupiters, with layout following Figure 2 of \citet{2018AJ....155..214T} and updated with NASA Exoplanet Archive confirmed planets.  The trend lines show the mass-radius relationship for equilibrium temperatures of 500, 1000, 1250, 1500, and 2000 K, assuming the mean composition and mean heating efficiency from the original figure.  HAT-P-67~b ($\star$ symbol) stands alone in a region defined by the lack of inflated sub-Saturns and explained by short lifetimes from runaway inflation and mass loss.}
    \label{fig:ThornFortUpdated}
\end{figure}

@@

The TESS Sector 26 lightcurve exhibits a weak $\sim$3 ppt peak-to-valley out-of-transit modulation. TESS Sectors 51-53 do not show as conspicuous a modulation signal but still show some ostensibly stellar variability.  The ebb and flow of the modulation can be seen in the minimally processed TESS lightcurve in Figure \ref{fig:TESSoverview}.  We fit the TESS lightcurve modulation with a quasiperiodic Gaussian Process (GP) model using \texttt{celerite} \citep{celerite1,celerite2}.  The GP model fit to the entire composite lightcurve yields a 4.7-day period; when fit to individual sectors alone the periods hover around 5.9 days.  The signal is weak enough, especially in Sectors 51-53, that \emph{TESS} instrumental systematics may contribute some of the out-of-transit modulation that we see in Figure \ref{fig:TESSoverview}.

We can independently constrain the expected stellar rotation period based on the stellar radius estimate, $v\sin{i_\star}$, and the observation of the planet's orbital inclination $i_p\sim90^\circ$ from orbit fitting and Doppler tomography \citep{2017AJ....153..211Z}.  We assume spin-orbit alignment, $i_\star \sim i_p$.  We adopt a high limit of $v\sin{i}=35.8\pm1.1$ km/s and a lower $v\sin{i}=30.9\pm2$ km/s value if macroturbulence is accounted for.  We obtain a range of $3.2 < P_\mathrm{rot}  < 4.8 $ days.  This range is typical for F stars that have not yet evolved too far into the subgiant branch \citep{2022ApJ...930....7A}.  The geometrical constraint comports with the 4.7-day GP-based modulation period derived from the stitched TESS lightcurve, but is lower than the 5.9-day per-Sector fits, slightly preferring the lower 4.7-day modulation as the stellar rotation period.  The 5.9-day value would require a stellar radius over 3.2$R_\odot$, which appears implausible.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/phase_2D_diagram_resid.png}
    \caption{Absorption depth phase scan over the entire HAT-P-67b orbit from 41 visits with HPF.  The planet orbital rest frame is shown as the black sine wave. Gaps in data at 10834-10836 \AA~ arise from telluric masking.  Unmasked telluric lines are perceptible at 10825 \AA.}
    \label{fig:HPFscanResid}
\end{figure}
